### 2015-06-17 chapters/ipa_meets_unicode.tex
update profile, first attempt at large table

\begin{itemize}
	\item Basic Latin (27 characters): \newline < a~b~c~d~e~f~h~i~j~k~l~m~n~o~p~q~r~s~t~u~v~w~x~y~z~.~| >
	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩
	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
	\item Greek and Coptic (3 characters): \newline < β θ χ >
%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
\end{itemize}

@@

There turn out to be a lot of characters in the IPA that could be conceived as 
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all 
such characters exist as well as precomposed combination in Unicode, and these 
precomposed characters should preferably be used. When combinations of a base 
character with diacritic are used, then these combinations are not canonical 
equivalent to the precomposed combinations. This means that any search will not 
find both at the same time.

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions. The conventions are influenced by the linguistic literature, though
some ad-hoc decisions had to be taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] rhotic hook <{\large \ ˞} >
	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
	  \item[→] labialized <ʷ>
	  \item[→] palatalized <ʲ>
	  \item[→] velarized <ˠ>
	  \item[→] pharyngealized <ˤ>
	  \item[→] glottalized <ˀ>
	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
	  \item[→] long <ː> \textbar{} half-long <ˑ>
	\end{itemize}
\end{itemize}

If a character sequence contains more than one diacritic below the base
character, then the place features are applied first (linguolabial, dental,
apical, laminal, advanced, retracted), followed by the manner features (raised,
lowered, advanced tongue root, retracted tongue root, frictionalized), then
secondary articulations (more round, less round, derhoticized), laryngeal
settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and
finally the syllabic or non-syllabic marker (for vowels, advanced and retracted
tongue root markers gets applied between the place and laryngeal setting). So:

@@

Finally, if a character sequence contains more than one diacritic above the base
character, we propose the following order, with Spacing Modifiers and tone
letters being added at the end:

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
	  \item[→] extra short <ĕ>
%	  \item[→] (tone accents, e.g. <è>)
	  \item[→] unreleased <e̚>
	  \item[→] Spacing Modifier Letters, see above
	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
	\end{itemize}	 
\end{itemize}

@@

Summary

>>>>

\begin{itemize}
	\item Basic Latin (27 characters): \newline < a~b~c~d~e~f~h~i~j~k~l~m~n~o~p~q~r~s~t~u~v~w~x~y~z~.~| >
	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
	\item Greek and Coptic (3 characters): \newline < β θ χ >
%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
\end{itemize}

@@

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used.\footnote{The IPA does not
describe any character for a voiced retroflex implosive, which would
transparently be \charis{ᶑ}. We propose to add this character to the IPA.} When
combinations of a base character with diacritic are used, then these
combinations are not canonical equivalent to the precomposed combinations. This
means that any search will not find both at the same time.

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions. The conventions are influenced by the linguistic literature, though
some ad-hoc decisions had to be taken.

If a character sequence contains more than one diacritic below the base
character, then the place features are applied first (linguolabial, dental,
apical, laminal, advanced, retracted), followed by the manner features (raised,
lowered, advanced tongue root, retracted tongue root, frictionalized), then
secondary articulations (more round, less round, derhoticized), laryngeal
settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and
finally the syllabic or non-syllabic marker (for vowels, advanced and retracted
tongue root markers gets applied between the place and laryngeal setting). So
the order that is proposed is the following (where <\textbar{}> indicates
\textit{or} and <→> indicates \textit{precedes}):

@@

Further, if a character sequence contains more than one diacritic above the base
character, we propose the following order. Note that this ordering is only
important when the extra short superscript is used. All other combinations will 
in practice never lead to problems.

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
	  \item[→] extra short <ĕ>
%	  \item[→] (tone accents, e.g. <è>)
	  \item[→] unreleased <p̚>
%	  \item[→] Spacing Modifier Letters, see above	  
 \end{itemize} \end{itemize}

Finally, when a character sequence contains more than one character in Spacing
Modifier Letters, these will be placed after all combining diacritic marks in the
following order:

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] rhotic hook <{\large \ ˞} >
	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
	  \item[→] labialized <ʷ>
	  \item[→] palatalized <ʲ>
	  \item[→] velarized <ˠ>
	  \item[→] pharyngealized <ˤ>
	  \item[→] glottalized <ˀ>
	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
	  \item[→] long <ː> \textbar{} half-long <ˑ>
	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
	\end{itemize}
\end{itemize}

@@

Summarizing the pitfalls discussed in this chapter, we propose to define a 
``strict'' IPA encoding to be used when interoperability of phonetic resources 
is intended. For regular research papers, when only the visual outline is 
important,

\begin{table}[htdp]

 \topcaption{IPA letters with Unicode encodings}
 \label{tab:ipa_letters}
 \tablehead{
   \toprule
   & Code & Unicode name & IPA name \\
  \midrule
  }
 \tabletail{\endline}

\begin{center}
\begin{xtabular}{ll>{\small}p{4cm}>{\small}l}
a & \uni{0061} & \textsc{latin small letter a} & open front unrounded \\ 
æ & \uni{00E6} & \textsc{latin small letter ae} & raised open front unrounded \\ 
ɐ & \uni{0250} & \textsc{latin small letter turned a} & lowered schwa \\ 
ɑ & \uni{0251} & \textsc{latin small letter alpha} & open back unrounded \\ 
ɒ & \uni{0252} & \textsc{latin small letter turned alpha} & open back rounded \\ 
b & \uni{0062} & \textsc{latin small letter b} & voiced bilabial plosive \\ 
ʙ & \uni{0299} & \textsc{latin letter small capital b} & voiced bilabial trill \\ 
ɓ & \uni{0253} & \textsc{latin small letter b with hook} & voiced bilabial implosive \\ 
c & \uni{0063} & \textsc{latin small letter c} & voiceless palatal plosive \\ 
ç & \uni{00E7} & \textsc{latin small letter c with cedilla} & voiceless palatal fricative \\ 
ɕ & \uni{0255} & \textsc{latin small letter c with curl} & voiceless alveolo-palatal fricative \\ 
d & \uni{0064} & \textsc{latin small letter d} & voiced alveolar plosive \\ 
ð & \uni{00F0} & \textsc{latin small letter eth} & voiced dental fricative \\ 
ɖ & \uni{0256} & \textsc{latin small letter d with tail} & voiced retroflex plosive \\ 
ɗ & \uni{0257} & \textsc{latin small letter d with hook} & voiced dental/alveolar implosive \\ 
\charis{ᶑ} & \uni{1D91} & \textsc{latin small letter d with hook and tail} & voiced retroflex implosive \\ 
e & \uni{0065} & \textsc{latin small letter e} & close-mid front unrounded \\ 
ə & \uni{0259} & \textsc{latin small letter schwa} & mid-central schwa \\ 
ɛ & \uni{025B} & \textsc{latin small letter open e} & open-mid front unrounded \\ 
ɘ & \uni{0258} & \textsc{latin small letter reversed e} & close-mid central unrounded \\ 
ɜ & \uni{025C} & \textsc{latin small letter reversed open e} & open-mid central unrounded \\ 
ɞ & \uni{025E} & \textsc{latin small letter closed reversed open e} & open-mid central rounded \\ 
ɤ & \uni{0264} & \textsc{latin small letter rams horn} & close-mid back unrounded \\ 
f & \uni{0066} & \textsc{latin small letter f} & voiceless labiodental fricative \\ 
ɡ & \uni{0261} & \textsc{latin small letter script g} & voiced velar plosive \\ 
ɢ & \uni{0262} & \textsc{latin letter small capital g} & voiced uvular plosive \\ 
ɠ & \uni{0260} & \textsc{latin small letter g with hook} & voiced velar implosive \\ 
ʛ & \uni{029B} & \textsc{latin letter small capital g with hook} & voiced uvular implosive \\ 
ɣ & \uni{0263} & \textsc{latin small letter gamma} & voiced velar fricative \\ 
h & \uni{0068} & \textsc{latin small letter h} & voiceless glottal fricative \\ 
ħ & \uni{0127} & \textsc{latin small letter h with stroke} & voiceless pharyngeal fricative \\ 
ʜ & \uni{029C} & \textsc{latin letter small capital h} & voiceless epiglottal fricative \\ 
ɦ & \uni{0266} & \textsc{latin small letter h with hook} & voiced glottal fricative \\ 
ɧ & \uni{0267} & \textsc{latin small letter heng with hook} & voiceless postalveolar+velar fricative \\ 
i & \uni{0069} & \textsc{latin small letter i} & close front unrounded \\ 
ɪ & \uni{026A} & \textsc{latin letter small capital i} & lax close front unrounded \\ 
ɨ & \uni{0268} & \textsc{latin small letter i with stroke} & close central unrounded \\ 
j & \uni{006A} & \textsc{latin small letter j} & voiced palatal approximant \\ 
ʝ & \uni{029D} & \textsc{latin small letter j with crossed tail} & voiced palatal fricative \\ 
ɟ & \uni{025F} & \textsc{latin small letter dotless j with stroke} & voiced palatal plosive \\ 
ʄ & \uni{0284} & \textsc{latin small letter dotless j with stroke and hook} & voiced palatal implosive \\ 
k & \uni{006B} & \textsc{latin small letter k} & voiceless velar plosive \\ 
l & \uni{006C} & \textsc{latin small letter l} & voiced alveolar lateral approximant \\ 
ʟ & \uni{029F} & \textsc{latin letter small capital l} & voiced velar lateral approximant \\ 
ɫ & \uni{026B} & \textsc{latin small letter l with middle tilde} & voiced pharyngealized alveolar lateral approximant \\ 
ɬ & \uni{026C} & \textsc{latin small letter l with belt} & voiceless alveolar lateral fricative \\ 
ɭ & \uni{026D} & \textsc{latin small letter l with retroflex hook} & voiced retroflex lateral approximant \\ 
ɮ & \uni{026E} & \textsc{latin small letter lezh} & voiced alveolar lateral fricative \\ 
ʎ & \uni{028E} & \textsc{latin small letter turned y} & voiced palatal lateral approximant \\ 
m & \uni{006D} & \textsc{latin small letter m} & voiced bilabial nasal \\ 
ɱ & \uni{0271} & \textsc{latin small letter m with hook} & voiced labiodental nasal \\ 
n & \uni{006E} & \textsc{latin small letter n} & voiced alveolar nasal \\ 
ɴ & \uni{0274} & \textsc{latin letter small capital n} & voiced uvular nasal \\ 
ɲ & \uni{0272} & \textsc{latin small letter n with left hook} & voiced palatal nasal \\ 
ɳ & \uni{0273} & \textsc{latin small letter n with retroflex hook} & voiced retroflex nasal \\ 
ŋ & \uni{014B} & \textsc{latin small letter eng} & voiced velar nasal \\ 
o & \uni{006F} & \textsc{latin small letter o} & close-mid back rounded \\ 
ø & \uni{00F8} & \textsc{latin small letter o with stroke} & close-mid front rounded \\ 
œ & \uni{0153} & \textsc{latin small ligature oe} & open-mid front rounded \\ 
ɶ & \uni{0276} & \textsc{latin letter small capital oe} & open front rounded \\ 
ɔ & \uni{0254} & \textsc{latin small letter open o} & open-mid back rounded \\ 
ɵ & \uni{0275} & \textsc{latin small letter barred o} & close-mid central rounded \\ 
p & \uni{0070} & \textsc{latin small letter p} & voiceless bilabial plosive \\ 
ɸ & \uni{0278} & \textsc{latin small letter phi} & voiceless bilabial fricative \\ 
q & \uni{0071} & \textsc{latin small letter q} & voiceless uvular plosive \\ 
r & \uni{0072} & \textsc{latin small letter r} & voiced alveolar trill \\ 
ʀ & \uni{0280} & \textsc{latin letter small capital r} & voiced uvular trill \\ 
ɹ & \uni{0279} & \textsc{latin small letter turned r} & voiced alveolar approximant \\ 
ɺ & \uni{027A} & \textsc{latin small letter turned r with long leg} & voiced alveolar lateral flap \\ 
ɻ & \uni{027B} & \textsc{latin small letter turned r with hook} & voiced retroflex approximant \\ 
ɽ & \uni{027D} & \textsc{latin small letter r with tail} & voiced retroflex tap \\ 
ɾ & \uni{027E} & \textsc{latin small letter r with fishhook} & voiced alveolar tap \\ 
ʁ & \uni{0281} & \textsc{latin letter small capital inverted r} & voiced uvular fricative \\ 
s & \uni{0073} & \textsc{latin small letter s} & voiceless alveolar fricative \\ 
ʂ & \uni{0282} & \textsc{latin small letter s with hook} & voiceless retroflex fricative \\ 
ʃ & \uni{0283} & \textsc{latin small letter esh} & voiceless postalveolar fricative \\ 
t & \uni{0074} & \textsc{latin small letter t} & voiceless alveolar plosive \\ 
ʈ & \uni{0288} & \textsc{latin small letter t with retroflex hook} & voiceless retroflex plosive \\ 
u & \uni{0075} & \textsc{latin small letter u} & close back rounded \\ 
ʉ & \uni{0289} & \textsc{latin small letter u bar} & close central rounded \\ 
ɥ & \uni{0265} & \textsc{latin small letter turned h} & voiced labial-palatal approximant \\ 
ɯ & \uni{026F} & \textsc{latin small letter turned m} & close back unrounded \\ 
ɰ & \uni{0270} & \textsc{latin small letter turned m with long leg} & voiced velar approximant \\ 
ʊ & \uni{028A} & \textsc{latin small letter upsilon} & lax close back rounded \\ 
v & \uni{0076} & \textsc{latin small letter v} & voiced labiodental fricative \\ 
ʋ & \uni{028B} & \textsc{latin small letter v with hook} & voiced labiodental approximant \\ 
ⱱ & \uni{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
ʌ & \uni{028C} & \textsc{latin small letter turned v} & open-mid back unrounded \\ 
w & \uni{0077} & \textsc{latin small letter w} & voiced labial-velar approximant \\ 
ʍ & \uni{028D} & \textsc{latin small letter turned w} & voiceless labial-velar fricative \\ 
x & \uni{0078} & \textsc{latin small letter x} & voiceless velar fricative \\ 
y & \uni{0079} & \textsc{latin small letter y} & close front rounded \\ 
ʏ & \uni{028F} & \textsc{latin letter small capital y} & lax close front rounded \\ 
z & \uni{007A} & \textsc{latin small letter z} & voiced alveolar fricative \\ 
ʐ & \uni{0290} & \textsc{latin small letter z with retroflex hook} & voiced retroflex fricative \\ 
ʑ & \uni{0291} & \textsc{latin small letter z with curl} & voiced alveolo-palatal fricative \\ 
ʒ & \uni{0292} & \textsc{latin small letter ezh} & voiced postalveolar fricative \\ 
ʔ & \uni{0294} & \textsc{latin letter glottal stop} & voiceless glottal plosive \\ 
ʕ & \uni{0295} & \textsc{latin letter pharyngeal voiced fricative} & voiced pharyngeal fricative \\ 
ʡ & \uni{02A1} & \textsc{latin letter glottal stop with stroke} & epiglottal plosive \\ 
ʢ & \uni{02A2} & \textsc{latin letter reversed glottal stop with stroke} & voiced epiglottal fricative \\ 
ǀ & \uni{01C0} & \textsc{latin letter dental click} & voiceless dental click \\ 
ǁ & \uni{01C1} & \textsc{latin letter lateral click} & voiceless alveolar lateral click \\ 
ǂ & \uni{01C2} & \textsc{latin letter alveolar click} & voiceless palatoalveolar click \\ 
ǃ & \uni{01C3} & \textsc{latin letter retroflex click} & voiceless (post)alveolar click \\ 
ʘ & \uni{0298} & \textsc{latin letter bilabial click} & voiceless bilabial click \\ 
β & \uni{03B2} & \textsc{greek small letter beta} & voiced bilabial fricative \\ 
θ & \uni{03B8} & \textsc{greek small letter theta} & voiceless dental fricative \\ 
χ & \uni{03C7} & \textsc{greek small letter chi} & voiceless uvular fricative \\
\end{xtabular}
\end{center}
\end{table}
